---
title: 'Motor trends: The effect of transmission type on fuel efficiency'
author: "John Lee"
date: "20 October 2015"
output: pdf_document
---
criteria:
Did the student interpret the coefficients correctly
Did the student do some exploratory data analysis
Did the student fit multiple models and detail their strategy for model selection
Did the student answer the question or detail why the question was unanswerable
Did the student do a residual plot and some diagnostics
Did the student quantify the uncertainty in their conclusions and or perform an inference correctly
Was the report brief
Did the report include an executive summary
Was the report done in rmd


#Executive summary paragraph

#main report (two pages only)
You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

    “Is an automatic or manual transmission better for MPG”
    "Quantify the MPG difference between automatic and manual transmissions"
    
```{r, results='hide'}
library(dplyr)
library(ggplot2)
library(car)
library(GGally)
data(mtcars)
mtcars$am <- factor(as.character(mtcars$am),labels = c("Automatic","Manual"))
```

The purpose of this analysis is to assess the effect of transmission type on the MPG of various cars. 

A basic violin plot indicates that without controlling for any other variables manual transmissions are typically more fuel efficent and getting more miles per gallon:
```{r}
ggplot(mtcars, aes(am,mpg)) +geom_violin() +geom_point() + xlab("Transmission Type")

```
A clear effect can be observed in the above plot. 
```{r}
summary(lm(mpg~ am,data = mtcars))$coef
```
Using tranmission type as predictors for the mpg variable we see that the mean mpg of automatic cars is 17.15 with a standard error of 1.1 and the average manual transmission is 7.2 mpg higher with a standard error of 1.8. A hypothesis test for the difference between the two categories yields a p-value < 0.001 suggesting that the transmission type strongly affects mpg. But is this the full story? We assess whether there may be confounding variables that are influencing this observation. A model using all other variables as predictors suffers from variance inflation but indicates some of the variables that may be useful in predicting the mpg of the various cars. Indeed correlation plots using these variables demonstrate extensive correlation between many of these variables 

```{r}
fit_all <- lm(data = mtcars, mpg~ .)
summary(fit_all)
vif(fit_all)
```
Weight is inversely correlated with mpg and captures 75% of the variation in this variable. A scatterplot of these two variables with the superimposed regression line suggests that a linear relationship may not be the most accurate method of relating these two variables.
```{r}
fit_wt <- lm(data = mtcars, mpg ~ wt)
summary(fit_wt)[c( "coefficients","r.squared")]
ggplot(mtcars, aes(mpg,x = wt)) + geom_point() + geom_smooth(method = "lm")   
```
This observation is supported by
```{r}
par(mfrow = c(2, 2))
plot(fit_wt)

```

```{r,ref.label= "predictors that capture much variance"}
fit_all <- lm(data = mtcars, mpg~ wt + hp + am + qsec)
summary(fit_all)
vif(fit_all)
```





A number of details of the engine (number of carburettors, number of cylinders and displacement ) are highly correlated with the power of the engine (see supplimentary figure 1). To simplify our analysis we only used the engine displacement as a measure of power fir each car. This variable is inversely correlated with mpg. A number of the other measures also showed correlations with the vehicle weight showing the strongest correlation:
 
```{r, cache=TRUE}
ggpairs(select(mtcars, mpg,disp, drat,wt,qsec,vs,gear))
ggpairs(select(mtcars, mpg,disp, drat,wt,qsec,vs,gear,am))
```


In order to attain a parsimonious model we assessed whether transmission type still affected mpg when controlling for just weight and engine displacement:
```{r}
fit <- lm(mpg~ wt + am, data = mtcars)
summary(fit)
vif(fit)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

#Appendix
can only include figures
```{r, cache = TRUE}
ggpairs(select(mtcars,mpg,cyl,disp,hp,carb,am), title = "Supp. Figure 1: Measures of engine output")

ggpairs(select(mtcars,wt,cyl,disp,hp,carb), title = "Supp. Figure 2: Vehicle weight")

ggpairs(select(mtcars, mpg, wt, hp, qsec,  am), title = "Supp. Figure 3")


```

```{r}
fit1 <- lm(data = mtcars, mpg~ disp)
fit2 <- update(fit1, mpg~ disp + carb)
fit4 <- update(fit1, mpg~ disp + carb + cyl)
fit3 <- update(fit1, mpg~ disp + carb + cyl+ hp)
anova(fit1,fit2,fit3,fit4)

```

